Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models

### 2015-03-26 tex/parametrized-learning.tex
reworked other work section

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
We demonstrate how discriminative classifiers can be used to approximate 
(generalized) likelihood ratio tests over high-dimensional data when a generative model 
for the data is available for training and calibration.  
\end{abstract}
%
%\begin{keywords}
%  Bayesian Networks, Mixture Models, Chow-Liu Trees
%\end{keywords}

@@

In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
Directly constructing the likelihood ratio for high-dimensional observations 
is often not possible or is computationally impractical. Here we demonstrate how 
discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for calibration.  We use the following notation

@@

Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.

@@

However, it is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
\begin{eqnarray}
\frac{f(x|\theta_0)}{f(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
\\
%&=& \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)}{ w_c(\theta_0) f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ f_{c'}(x| \theta_1)}{  f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ g_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ g_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.  The third line uses the previous results to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).

@@

In \cite{ClaytonScott,JMLR:v14:tong13a}, the authors consider the machine learning problem associated to Neyman-Pearson hypothesis testing. As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d. samples from each hypothesis. This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification. Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
In \citep{TommiJaakkola}, the authors explore a way of leveraging generative models to derive kernel functions for use in discriminative methods. This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration. In \citep{BiancaZadrozny}, the authors emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this. In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests.

Mention also \citep{McCallum}

>>>>

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
We demonstrate how discriminative classifiers can be used to approximate 
generalized likelihood ratio tests over high-dimensional data when a generative model 
for the data is available for training and calibration.  
\end{abstract}
%
%\begin{keywords}
%  Bayesian Networks, Mixture Models, Chow-Liu Trees
%\end{keywords}

@@

In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
Directly constructing the likelihood ratio for high-dimensional observations 
is often not possible or is computationally impractical. Here we demonstrate how 
discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.  We use the following notation

@@

Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.  Similar points are made in~\citep{ClaytonScott}.

@@

However, it is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
\begin{eqnarray}
\frac{f(x|\theta_0)}{f(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
\\
%&=& \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)}{ w_c(\theta_0) f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ f_{c'}(x| \theta_1)}{  f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ g_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ g_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.  The third line uses Theorem 1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).

@@

\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing. As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d. samples from each hypothesis. This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification. Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
\cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods. This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.  \cite{McCallum}
 consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.  \cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this. In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests. \cite{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.
